Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-08-30 TACO-paper-v2/coffee.tex
Add few lines, but almost time to go out

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our optimizations for cross-loop arithmetic intensity are actually effective and that in a field such as automated code generation for finite element methods they can significantly impact full-application performance. The next, logical question is about whether applications in other computational domains would benefit from the same transformations that COFFEE is capable of applying; or, more in general, to what extent a general-purpose compiler should (could) integrate some of them.

We begin observing that, strategically, our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This is because there are neither conceptual restrictions nor technical reasons for which the compiler could not be integrated in frameworks other than Firedrake. It would be quite easy, for example, to achieve that with the popular FEniCS framework, the pioneer of automated code generation for finite element methods. More challenging is to assess the generality of the approach; that is, if and how COFFEE and its optimizations could be transferred to other computational domains, and to what extent that is really useful. To answer this question, we need to jump back to the origins of the compiler.

@@

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ computing the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives. Translating into efficient code such a formulation of local assembly is a challenging process, especially for complicated equations; this was actually one of the problems motivating research in DSLs and automated code generation. Premised that, and abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
\begin{enumerate}
\item Expression manipulation at the mathematical layer (e.g. simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality. The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. Whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.  
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation is about resource-aware execution. It is indeed true that, in the context of COFFEE, it has been successfully applied to improve register pressure; however, the general idea is to re-schedule (re-associate) operations so as to maximize some parameters. Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel. An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlo-lcpc}. 
\item The quantities in the local assembly expression are known at code generation time. This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
\item TODO lack of stencils -> simplifies most of the proposed code transformations; the interplay with a ployhedral/general-purpose compiler would need to be studied to understand how the transformations for arithmetic intensity would impact things like auto-vectorization.
\item TODO: cross-loop vectorization ... relationship between indices/memory access patterns and vectorization techniques.
\end{enumerate}

From the perspective of the domain-specific language and compiler designers, our technique represents an incentive to produce an extremely simple representation of the code (e.g. inlined mathematical expressions and perfect loop nests, if possible, in the form of an abstract syntax tree in the case of COFFEE) and then leave the (possibly domain-aware) code optimizer responsible for choosing and applying the best set of transformations for the underlying architecture. This aspect is of crucial importance since it allows achieving \textit{separation of concerns}, with domain specialists that can completely abstract from the performance optimization problem.

>>>>

\section{Generality of the Approach and Applicability to Other Domains}
\label{sec:generality}
We have demonstrated that our optimizations for cross-loop arithmetic intensity are effective in the context of automated code generation for finite element local assembly. In this section, we discuss about the potential for their applicability in other computational domains and, more in general, to what extent a general-purpose compiler could integrate some of them.

We begin observing that our strategic choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This is because there are neither conceptual nor technical reasons for which the compiler could not be used to optimize codes different than those produced by Firedrake. It would be quite easy, for example, to integrate COFFEE with the popular FEniCS framework, the pioneer of automated code generation for finite element local assembly. More challenging is to assess the generality of our optimization strategy; that is, if and how COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent it would be helpful for improving full-application performance. To answer these questions, we need to go back to the origins of our compiler.

@@

The expression represents the numerical evaluation of an integral at $n_1$ points in the mesh element $K$ to compute the local element matrix $A$. Functions $\alpha$, $\beta$ and $\gamma$ are problem-specific and can be intricately complex, involving for example the evaluation of derivatives. Abstracting from the inherent structure of $\alpha$, $\beta$ and $\gamma$, a number of aspects can be highlighted
\begin{enumerate}
\item Expression manipulation at the mathematical layer (e.g. simplification, decomposition into sub-expressions) opens multiple semantically-equivalent code generation opportunities, characterized by different trade-offs in parallelism, redundant computation, and data locality. The basic idea is to exploit properties of arithmetic operators, like associativity and commutativity, to re-schedule the computation suitably for the underlying architecture. Loop-invariant code motion and expression splitting follow this principle, so they can be re-adapted or extended to any domain computationally similar to local assembly, like those involving tensor contractions~\cite{TCE}.
\item In the expression above, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. The various sub-expressions, such as $a$, $b$, and $c$, in general only depend on (i.e. iterate along) a subset of the enclosing loops. In addition, some of these sub-expressions might reduce to the same values as iterating along certain iteration spaces. This code structure motivated the generalized loop-invariant code motion technique. Whenever sub-expressions invariant with respect to different sets of affine loops can be identified, the question of whether, where and how to hoist them, while minimizing redundant computation, arises.  
\item Associative arithmetic operators are the prerequisite for expression splitting. In essence, this transformation is about resource-aware execution. It is indeed true that, in the context of COFFEE, it has been successfully applied to improve register pressure; however, the general idea is to re-schedule (re-associate) operations so as to maximize some parameters. Expression splitting could be used, for example, as a starting point to perform kernel fission; that is, dividing a kernel into multiple parts, each part characterized by memory requirements less stringent than the original kernel. An example of this, specifically about fissioning associative non-affine loops in unstructured mesh applications, is~\cite{carlolcpc}. 
\item The quantities in the local assembly expression are known at code generation time. This means that ``good'' code can be generated, for example with explicit loop bounds (i.e. no parameters), statically-initialized constant n-dimensional arrays, and no pointer aliasing. All of these factors contribute to the effectiveness of source-to-source code transformations: by knowing iteration spaces in advance and having a clean memory access pattern, transformations such as padding, data alignment, and vector-register tiling can be automated by a compiler like COFFEE.
\item Compiler auto-vectorization has become effective for a variety of codes. Nevertheless, multi-loop vectorization, which involves loading and storing data along a subset of the loops characterizing the iteration space (rather than just along the innermost loop), is not supported by available general-purpose and polyhedral compilers to the best of our knowledge. However, the outer-product vectorization technique presented in this paper shows that two-loop vectorization can outperform standard auto-vectorization; more importantly, we expect the performance gain to scale with the number of vectorized loops and the vector length (as proved by the Xeon Phi results). Despite being not straightforward to automate multi-loop vectorization in a general-purpose compiler, especially if stencils are present, we believe that in specific domains this could be more easily achieved. The intuition is to map the memory access pattern on vector registers, and then relying on in-register shuffling to minimize the traffic between memory and processor.
\end{enumerate}

We conclude highlighting the fact that, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g. fully-inlined mathematical expressions in the form of an abstract syntax tree) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations. This aspect is of crucial importance since it allows achieving \textit{separation of concerns}, with domain specialists that can completely abstract from the performance optimization problem.
